---
title: "Final Project: Himalayan Bird Abundance Data"
date: "Due 3:00 PM 7/29/21"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
```

## Himalayan Data Collected by Dr. Alex White

Watch the video introduction by Dr. Alex White, "Himalayan Data Intro" on GauchoSpace.

## The data

Two data tables are provided for the final project. The first table 1) himalayan_bird_abundance.csv was collected by Dr. Alex White as part of his dissertation. The second table 2) himalayan_forest_metadata.csv was compiled by Dr. Alex White for each sample site using historic bio climatic data from the https://worldclim.org/data/bioclim.html. Both data tables can be found in the data folder of this project. Let's read each table in look at the data we have to work with.

Let's start with the birds.

```{r,message = FALSE}
birds <- read_csv("/cloud/project/data/himalayan_bird_abundances.csv")
dim(birds)
```

It looks like there are 38 observations (rows) and 305 variables (columns) in the birds dataset.  
Next, let us look at the bioclimatic data for each sampling site.

```{r,message = FALSE}
forest_metadata <- read_csv("/cloud/project/data/himalayan_forest_metadata.csv")
dim(forest_metadata)
```
For this table we have 38 observations (rows) and 9 variables (columns) in the forest_metadata.
```{r}
birds_gather <- birds %>% 
  gather(key="species", value="n",-X1)
data_joined_1 <- inner_join(forest_metadata,birds_gather,by=c('grid'='X1')) %>% 
  filter(n>0)
```

Here is some information that will help you understand the 9 variables:

1.	grid = site location of each plot
2.	elevation = meters with decimal
3.	lat = latitude in decimal degrees (dd)
4.	long = longitude in decimal degrees (dd)
5.	region = “E” for East and “W” for West region of Himalayas
6.	max_temp_of_warmest_month, note: (Celsius no decimal, e.g. “300”= 30.0 C)
7.	temp_annual_range, note: (unit in Celsius, does not include decimal, thus “300”= 30.0 C)
8.	annual_precip = Annual Precipitation, note: (unit in millimeters)
9.	precip_seasonality = Precipitation Seasonality (Coefficient of Variation), note: How much rainfall changes from month to month. When the coefficient is high, the rainfall changes a lot from month to month so there is a high amount of seasonality. 

## Final Project Questions!
Please share visualizations where you think it is necessary.

```

1.	Review the data table’s structure and describe any key elements and initial challenges your research team encountered. 
    a)	Share your unique solutions to the challenges
    b)	What information is shared between each table? 
    c)	Would you recommend any tidydata best practices for either table? How does these tables represent good tidydata practices?
    
2.	Explain the relationship between the data frames and how they may be linked?   
  
3.	Create a data frame. Compare each region (East and West) in your new dataframe by using the variables that you find most useful for answering the following 3 questions:  
    a) How many individual birds were sighted in each grid?  How many individual birds were sighted overall in each region?   
```{r}
birds_in_grid <- data_joined_1 %>%
  group_by(grid) %>% 
  summarize(n_birds=sum(n))

birds_in_region <- data_joined_1 %>%
  group_by(region) %>% 
  summarize(n_birds=sum(n))

```
    
    b) How many bird species were present in each grid?  How many bird species were present overall in each region?
```{r}
species_in_grid <- data_joined_1 %>%
  group_by(grid) %>% 
  summarize(n_species=n_distinct(species))

species_in_region <- data_joined_1 %>%
  group_by(region) %>% 
  summarize(n_species=n_distinct(species))

```
    
    c) Which grid had the highest number of bird species in each region? What is the elevation and precipitation of each of those grid locations? Name the three most common bird species in each of those grid locations? 
```{r}
data_joined_1 %>%
  group_by(grid,region) %>% 
  summarize(n_species=n_distinct(species)) %>% 
  arrange(desc(n_species)) %>% 
  head(n=2)

data_joined_1 %>%
  filter(grid=="A4"|grid=="U1") %>% 
  group_by(grid,elevation,annual_precip) %>% 
  summarize(n_species=n_distinct(species))

max_species_by_n <- data_joined_1 %>%
  filter(grid=="A4"|grid=="U1") %>%
  select(grid,region,species,n) %>% 
  group_by(grid) %>%
  slice_max(n=3,order_by=n)

```

    d) Within each region, which grid location had the least amount of bird species observed?  Which species were observed in each of those grid locations? Visualize all the elevations where those bird species are found.  Separate this by region. Explain the trends that you see on the graph (Hint: use a line graph and color by species). 
```{r}
data_joined_1 %>%
  filter(species=="Tarsiger_chrysaeus"|
         species=="Cettia_brunnifrons"|
         species=="Trochalopteron_affine"|
         species=="Carpodacus_thura"|
         species=="Prunella_strophiata"|
         species=="Tarsiger_chrysaeus"|
         species=="Phoenicurus_frontalis"|
         species=="Turdus_maximus"|
         species=="Phylloscopus_pulcher"|
         species=="Leucosticte_nemoricola"|
         species=="Anthus_roseatus") %>% 
  ggplot(mapping=aes(x=n,y=elevation,color=species))+
  geom_line()+
  facet_grid(facets=vars(region))

```


4.	The Himalayan landscape is known for its significant changes in elevation. Let’s organize the data to show the total number of species across the elevational gradient. What about the total number of individual birds across the same elevational gradient?  
    a)	When you compare and contrast your analysis for the west vs. east do you notice any trends? How does the metadata inform the trends that you see? What theories do you have that might provide an explanation for what might be occurring?
```{r}
#Total species across elevational gradient
data_joined_1 %>%
  group_by(elevation) %>% 
  summarise(n_species=n_distinct(species))

#Total number of individual birds across elevation gradient
data_joined_1 %>%
  group_by(elevation)%>%
 	 summarise(n_individual_birds=sum(n))
 	 
#Line Graph of # of species in each region across elevation gradient
data_joined_1 %>%
  group_by(elevation,region) %>% 
  summarise(n_species=n_distinct(species)) %>% 
  ggplot(mapping=aes(x=elevation,y=n_species,color=region))+
  geom_line()
  
#Line Graph of # of individual birds in each region across elevation gradient
data_joined_1 %>%
  group_by(elevation,region) %>% 
  summarise(n_individuals=sum(n)) %>% 
  ggplot(mapping=aes(x=elevation,y=n_individuals,color=region))+
  geom_line()


```


    b)	Using these online resources (and any others you find), please share any life history characteristics that might provide insight on the birds’ status (for example, conservation status, habitat or food preference, competition, etc.)     
        i.	Resources:   
            1.	Avibase: https://avibase.bsc-eoc.org  
             2.	Birds of the World: https://birdsoftheworld.org  
              3.	Birdlife India: http://datazone.birdlife.org/country/india      

5.	Bioclimatic data was provided for each site. These include max temp for warmest month, temp annual range, annual precipitation, and/or precipitation seasonality.   
    a)	What climate data did you find most insightful to interpreting the bird data? Please explain.        
```{r}
#shows the relationship between max_temp_of_warmest_month and how many species there are
data_joined_1 %>%
  group_by(max_temp_of_warmest_month, region) %>%
  summarize(n_species = n_distinct(species)) %>%
  ggplot(mapping=aes(x=max_temp_of_warmest_month,y=n_species,color=region))+
  geom_point()


#shows the relationship between max_temp_of_warmest_month and how many total birds there are
data_joined_1 %>%
  group_by(max_temp_of_warmest_month, region) %>%
  summarize(n_individuals=sum(n)) %>%
  ggplot(mapping=aes(x=max_temp_of_warmest_month,y=n_individuals,color=region))+
  geom_point()
  
#shows the relationship between temp_annual_range and how many species there are
data_joined_1 %>%
  group_by(temp_annual_range, region) %>%
  summarize(n_species = n_distinct(species)) %>%
  ggplot(mapping=aes(x=temp_annual_range,y=n_species,color=region))+
  geom_point()

#shows the relationship between temp_annual_range and how many total birds there are
data_joined_1 %>%
  group_by(temp_annual_range, region) %>%
  summarize(n_individuals=sum(n)) %>%
  ggplot(mapping=aes(x=temp_annual_range,y=n_individuals,color=region))+
  geom_point()

#shows the relationship between annual_precip and how many species there are
data_joined_1 %>%
  group_by(annual_precip, region) %>%
  summarize(n_individuals=sum(n)) %>%
  ggplot(mapping=aes(x=annual_precip,y=n_individuals,color=region))+
  geom_point()

#shows the relationship between precip_seasonality and how many species there are
data_joined_1 %>%
  group_by(precip_seasonality, region) %>%
  summarize(n_species = n_distinct(species)) %>%
  ggplot(mapping=aes(x=precip_seasonality,y=n_species,color=region))+
  geom_point()

#shows the relationship between precip_seasonality and how many total birds there are
data_joined_1 %>%
  group_by(precip_seasonality, region) %>%
  summarize(n_individuals=sum(n)) %>%
  ggplot(df,mapping=aes(x=precip_seasonality,y=n_individuals,color=region))+
  geom_point()

```

```{r}

data_joined_1 %>%
  group_by(region,max_temp_of_warmest_month) %>%
  summarize(n_species = n_distinct(species)) %>%
  ggplot(mapping=aes(x=max_temp_of_warmest_month,y=n_species,color=region))+
  geom_point(color = "black") +
  geom_smooth()+
  facet_grid(facets=vars(region))+
  labs(x= "Maximum Temperature of Warmest Month", y= "Number of Species")
  
#most biodiversity when it's hot for E?

data_joined_1 %>%
  group_by(region,temp_annual_range) %>%
  summarize(n_species = n_distinct(species)) %>%
  ggplot(mapping=aes(x=temp_annual_range,y=n_species,color=region))+
  geom_point(color = "black") +
  geom_smooth()+
  facet_grid(facets=vars(region))+
  labs(x= "Annual Range of Temperature", y= "Number of Species")
#E birds prefer even temperatures throughout the year?

data_joined_1 %>%
  group_by(region,annual_precip) %>%
  summarize(n_species = n_distinct(species)) %>%
  ggplot(mapping=aes(x=annual_precip,y=n_species,color=region))+
  geom_point(color = "black") +
  
  geom_smooth()+
  facet_grid(facets=vars(region))+
  labs(x= "Annual Precipitation", y= "Number of Species")+
  theme_dark()

data_joined_1 %>%
  group_by(region,precip_seasonality) %>%
  summarize(n_species = n_distinct(species)) %>%
  ggplot(mapping=aes(x=precip_seasonality,y=n_species,color=region))+
  geom_point(color = "black") +
  geom_smooth()+
  facet_grid(facets=vars(region))+
  labs(x= "Precipitation Seasonality", y= "Number of Species")




```


6. Choose your own adventure!  Present an additional visualization or trend in the data that you found useful in understanding the Himalayan region.

7.	Share what next steps you might take to investigate this data further.


## Ummm... What?
Don't worry! You are not expected to submit a 50-page report here. We want you to work together and contribute to this ongoing project, and we hope you will motivate each other and find ways to engage together on this group project. So far, we've spent these 5 weeks doing a lot of following along and working with data from projects we don't know much about. Here is your chance to be creative with your new skills that your instructor has worked on as part of his dissertation. Have fun!

## Items due before your discussion section 

**Here is what you need to bring to your next discussion section:**

1. Provide tidy data suggestions to improve the data tables. How does these tables represent good tidydata practices?

2. Bring your code to your discussion section and be ready to share and compare ideas.

3. Code (R chunk) graphs/visuals that answer the questions above. In lecture 5 we learned how to use ggplot to create visual representations of data. What kind of visual can you use to answer the questions above? There may be multiple, but in the end your group will choose visualizations you think best describe the data.

**Here is what you need to email your peer mentor ahead of your final discussion section:**

1. Draft outline of your video presentation. It should be about 10-15 minutes in length. You can use ZOOM to record your presentation. Here is a recommended outline:

    + Introduce your group.
    + Provide an overview of the data tables.
    + Provide a step-by-step solutions for each question, and don't forget to share your code.
    + Provide a stong closing statement that drills down on some key take aways.
    + Be prepared for Q and A.

### Here are some tips for your presentation

* Remember to keep it simple.
* Avoid jargon.
* Speak slowly and clearly.
* Use less text and more images/visuals on slides.
* Include your group: team member names, colleges/universities, and majors on the first slide.
* Make sure to thank your audience at the end - it also signals to the audience you are done.
 
### ZOOM Local Recording Instructions
 
Here is a helpful zoom link for creating these recordings (https://support.zoom.us/hc/en-us/articles/201362473-Local-recording). If you need any help, reach out to your peer mentor for additional support!
 
Zoom Local Recording Instructions:

1. Prerequisites:

  + Zoom Basic (Free) account or above
  + Zoom Desktop Client for Windows, macOS, or Linux, version 2.0 or higher (https://zoom.us/support/download)
  
2. Enabling local recording:

  + Sign into the Zoom web portal as an administrator with the privilege to edit account settings.
  + In the navigation menu, click Account Management then Account Settings.
  + In the Recording tab, navigate to the Local Recording option and verify that the setting is enabled.
  + If the setting is disabled, click the toggle to enable it. If a verification dialog displays, click Turn On to verify the change.

3. Starting a local recording:

  + The host must record the meeting or grant the ability to record to a participant.
  + Start a Zoom meeting as the host.
  + Click the option to Record.
  + If there is a menu, select Record on this Computer.
  + A recording indicator will appear in the top-left corner while recording is active.
  + After the meeting has ended, Zoom will convert the recording so you can access the files.
  + Once the conversion process is complete, the folder containing the recording files will open.
  + By default, the audio/video file (MP4) will be named Zoom_0.mp4. The audio only file (M4A) is named audio_only.m4a. **Please change this, include your name so we can tell the videos apart.**
  

### Instructions for uploading to Box

Note: The student assigned the task of uploading the final video file into Box must follow the steps below.

1. Please let your peer mentor know that you will be uploading the video for the group. Provide the email you will be using to upload.
2. You will receive an email invitation link to our "UCSB-Smithsonian Scholars ERES Final Project Videos 2021" Box account.

  + Accept the invitation
  + Create a Box account
  
3. You will have access to the Box folder, but only to upload. When you are ready you can upload the video by clicking "Upload" on the top right of the screen.

### Roles for the group project

Your project team consists of the members in your discussion section group. Please work as a team, and make sure to include everyone. Your peer mentor will be your guide through the project and will provide feedback during discussion section. It's important to come prepared. 

Part of this exercise is learning to work with others in a team setting. Think about which roles would be a good fit with your strengths. You may have more than one role.

Roles:

  * Presenter / Screen share
  * Slideshow maker
  * Coders 
  * Data Visualizers
  * Tidy Data Summarizer
  * Zoom Recorder
  * Zoom Uploader 

If you have any questions, please contact your peer mentor or the program coordinator. Happy coding!


...
